Iterative Monomorphisation

### 2025-03-03 iterative_monomorphisation.tex
fix
Mark Summerfield suggestions

Automatic theorem provers are employed to provide automation to users of proof assistants. Many proof assistants, such as HOL4~\cite{slind-norrish-2008}, HOL Light~\cite{harrison-2009} and Isabelle\slash HOL \cite{nipkow-et-al-2002}, support rank-1 polymorphism, where type quantification is allowed at the top level of formulae. By contrast, many automatic provers operate only on monomorphic logics. One approach to closing this gap is to extend provers to natively support polymorphism, as has been done for Vampire~\cite{bhayat-reger-2020}. This, however, entails a lot of work that needs to be redone for every prover.

@@

Any translation method based on instantiation is
bound to be incomplete.
By a type version of the compactness theorem, there exists for any first order polymorphic formula \(\varphi\) an equisatisfiable finite set of monomorphic instances of \(\varphi\). However, this set cannot be computed~\cite[Theorem~1]{expr-poly-types}.

B\"ohme's iterative approach is implemented as part of Isabelle/HOL's SMT (satisfiability modulo theories) integration~\cite[Chapter 2]{sb-phd}. This implementation is also used by Sledgehammer~\cite{judgement,paulson-blanchette-2010} to interface with superposition based automatic theorem provers. However, it is is documented only as a single subsection in B\"ohme's PhD thesis~\cite[Section 2.2.1]{sb-phd}.
A similar algorithm appears to be implemented in the \texttt{MESON} tactic~\cite{harrison-1996} of HOL Light, but it is undocumented.

In this paper, we present an algorithm based on our understanding of Böhme's description and implementation (Section~\ref{sec:high level-algorithm}). We also provide a more detailed description to help future implementers. In addition, our description shows some of the ways in which an implementation can be made more efficient and avoid explosions (Section~\ref{sec:low-level-algorithm}).

Briefly, the algorithm works as follows. We assume problems to be sets of formulae. All symbols of the problem are collected, and the polymorphic instances of symbols are matched against the monomorphic ones. This yields new symbol instances, both polymorphic and monomorphic. The process is then iterated, making use of the newly generated instances.

Consider the unary type constructor \ty{list}. If a formula contains $\ty{list}(\alpha)$, where $\alpha$ is a type variable, the types $\ty{list}(\ty{int})$, $\ty{list}(\ty{list}(\ty{int}))$, etc., can be generated. However, because new types emerge through matching, $\ty{list}(\ty{list}(\ty{int}))$ can be obtained only once the $\ty{list}(\ty{int})$ instance has been generated. This example also makes it clear that the set of instances that can be generated is infinite.

To keep the number of generated formulae finite, we limit the number of iterations. After the iterations are completed, the new monomorphic symbol instances are used to instantiate the polymorphic symbols in the problem's formulae, generating new monomorphic formulae. Finally, because monomorphic provers support only nullary type constructors, types must be `mangled'; for example, the compound type $\ty{list}(\ty{int})$ could be mangled to $\ty{list\_int}$.

@@

The algorithm is sound because the newly generated formulae are instances of the initial problem's formulae. %, where type variables have been instantiated with monomorphic types.
It is, however, not complete.

@@

This example illustrates how an infinite number of new formulae can be generated from a simple initial problem.
%Although the example may seem contrived, similar problems frequently arise in practice. For example, the \sym{concat} function of Isabelle/HOL, which is characterised in the base case by the equation \(\sym{concat}\langle\alpha\rangle\; \sym{Nil}\langle\ty{list}(\alpha)\rangle = \sym{Nil}\langle\alpha\rangle\), exhibits the same behavior.
Any reasonable implementation requires bounds limiting the number of new type arguments, substitutions and formulae.

@@

\looseness=-1
The algorithm presented above is too naive for an implementation. In this section, we present a lower level algorithm with the following features.
First, numeric bounds are introduced to stop explosive enumerations.
Second, type argument tuples are separated into an old set and a new set to avoid re-computing some of the same matchings in successive iterations.
Third, substitutions are directly applied to the type arguments instead of the formulae. This avoids having to re-extract the type arguments from the formulae at each iteration. New formulae are generated only once all iterations are complete, in a separate, final step.

The data structures used in the algorithm are based on the ones used in the high level description. Instead of a map \(M\) from symbols to monomorphic\pagebreak[2] type argument tuples, we now have \(M_\text{old}\) and \(M_\text{new}\), which play the same role whilst also distinguishing between those type argument tuples that have already been matched against and those that have not. Similarly, \(N_\text{old}\) and \(N_\text{new}\) replace the map \(N\) from formulae to symbols to non-monomorphic type argument tuples. Finally, maintain a map \(S\) from formulae to the substitutions generated by the matchings. It is used to generate new formulae in the final phase.

@@

The iterative monomorphisation algorithm is given in Figure~\ref{iter_mono}. It has three phases. The first phase applies a \emph{monomorphisation step} to each formula in \(\Phi\) until the user-set limit, \textcolor{ourblueviolet}{num\_loops}, is reached. This limit is the only bound necessary for the algorithm to terminate. We use the colour blue to identify bounds and code related to bounds. At the end of each of these iterations, the old and new type argument maps are updated with newly generated types. Once these iterations are completed, the first phase is done and the substitutions used to create new type argument tuples are passed to $\mathit{generate\_mono\_formulae}$ for the second phase. The third phase mangles the composite types of the newly monomorphised formulae. This allows targeting a simply typed logic with no support for $n$-ary type constructors.
%~\cite[Sections 4 and 5]{mono-trans}.
% Paper name is gt

@@

Similar bounds are used for the non-monomorphic type argument tuples:
The limit on the number of new non-monomorphic type argument tuples is \(\min(\max(\PolyMult\cdot n, \PolyFloor),\allowbreak \PolyCap)\), where \(n\) is the number of non-monomorphic type argument tuples associated with the current formula. An important difference with the monomorphic case is that \(n\) depends on the current formula being processed whilst \(m\) does not.
Both in the monomorphic and in the non-monomorphic case, the maximum number of newly generated type argument tuples is fixed per formula and iteration.

@@

                  \(\sigma\leftarrow\sigma_1\circ\dots\circ\sigma_n\) \phantom{\big S}

@@

The various bounds presented here overlap to some extent. For instance, having at most \textcolor{ourblueviolet}{substitution\_cap} substitutions generated by \emph{matches} may be sufficient to curb the number of new type argument tuples, making the \textcolor{ourblueviolet}{\MonoCap}, \textcolor{ourblueviolet}{\MonoMult}, \textcolor{ourblueviolet}{\MonoFloor} triplet superfluous. Each bound nonetheless has its uses. For example, problems that lead to few successful matches but many type argument tuples may benefit from a limit on the number of new type argument tuples whilst problems for which substitution generation is more explosive may benefit from a limit on the number of generated substitutions.

@@

The Zipperposition implementation of the \emph{mono\_substs} function uses a map from type variables to substitutions instead of a set to filter the relevant substitutions from \(S\) efficiently. The \textcolor{ourblueviolet}{max\_substs} bound exists for two main reasons:
\begin{enumerate}
   \item The iterative monomorphisation algorithm can generate up to \textcolor{ourblueviolet}{max\_substs} new monomorphic formulae per initial polymorphic formula. Generating an excessive number of new formulae can overwhelm the downstream prover. The final number of output formulae is limited to at most \(|\Phi|\cdot \text{\textcolor{ourblueviolet}{max\_substs}}\).
   \item The \emph{mono\_substs} function is the algorithm's most explosive part. For a formula \(\varphi\), if \(S(\varphi)\) contains \(n\) substitutions that each instantiate exactly one of \(v\) %different
type variables, up to \(n^v\) monomorphising substitutions may be generated. Recall that the total number of type argument tuples used to generate \(S(\varphi)\) can be doubly exponential in the number of loop iterations. The starting \(n\) may therefore already be very large.
\end{enumerate}
%[This function can be thought of as exploring the tree of all possible monomorphising substitutions, \(S_{\text{res}}\) would represent the leaves that have already been visited and \(\sigma\) would represent the current node of the tree.]

@@

Each bound of the monomorphisation process presents a tradeoff: a higher bound allows for a more exhaustive instantiation of type variables but takes more time. Since we cannot test all possible combinations of values for all bounds to find the best compromise between completeness and speed,
we group closely related bounds together and test combinations of values for the bounds in these groups. Once we find the best performing set of values for a group, we assign these values to the corresponding bounds as we begin the search for the next group. If several groups result in the same number of proved problems, we select the most constraining values. Winning entries are shown in bold in Tables \ref{mono_ty_args} to \ref{e_settings}.

@@

Table~\ref{mono_ty_args} groups bounds that control the maximum number of newly generated monomorphic type arguments per formula and iteration. The limit on newly generated type arguments is determined by three components that form a natural group of bounds. The table shows that generating no new monomorphic type arguments seems to be the best approach. This result may seem counterintuitive, but it is possible to monomorphise formulae without generating monomorphic type arguments. This is because non-monomorphic type argument generation can generate substitutions that instantiate one or more type variables, and it is these substitutions that are used to monomorphise formulae.

@@

\looseness=-1
Table~\ref{nmon_ty_args} plays the same role as Table~\ref{mono_ty_args} except that it evaluates the bounds limiting the number of new \emph{non-}monomorphic type arguments. The bound values are lower because we found non-monomorphic type arguments to be explosive in preliminary evaluations. The table confirms that non-monomorphic type argument generation drives the creation of useful non-monomorphic formulae. This is indicated by the very low number of problems solved when no new non-monomorphic type arguments are allowed. Performance of the monomorphisation algorithm seems to plateau for some ranges of values and drops off beyond.
%TODO check that 125 is what we get if we just run Zipperposition for 10 seconds (maybe with all additional problems that are easily monomorphised)

@@

For larger problems, the monomorphisation algorithm may time out despite the bounds. In these cases neither Zipperposition nor E will have had a chance to try to solve the problem. To avoid this, a timer can interrupt the monomorphisation algorithm, after which Zipperposition resumes normal operation. Table~\ref{mono_time} tests the amount of time that is allocated to monomorphisation against the number of iteration of the monomorphisation algorithm. Neither parameter seems to substantially affect the performance of the algorithm.

@@

Table \ref{e_settings} shows the impact of the options with which we call the E prover. The call point corresponds to the fraction of the total timeout of Zipperposition's timeout after which the normal operation of Zipperposition is interrupted to begin the monomorphisation process. The E timeout limits for how long E is run before being interrupted, at which point Zipperposition resumes normal operation. The longer Zipperposition runs before E is invoked, the more formulae are generated, and the more explosive iterative monomorphisation is. This likely explains the poor performance for greater values of the E call point.

@@

To evaluate the usefulness of iterative monomorphisation as an alternative to native polymorphism, two competitive higher order polymorphic provers were run on the 1034 problem suite. We chose Leo-III and Zipperposition. We unfortunately had to exclude Vampire because of parsing issues.

@@

\looseness=-1
We see the following avenues for future work. First, iterative monomorphisation
blindly enumerates candidate instantiations, without exploiting any knowledge
about the logical structure of the formulae in which symbols occur. For
example, a lemma $\sym{p}\langle\alpha\rangle$ cannot be used to prove the
conjecture $\lnot \sym{p}\langle\ty{nat}\rangle$ because of the incompatible
polarities, but our algorithm instantiates $\alpha$ with $\ty{nat}$ regardless.
Second, some automatic provers as well as tools such as Sledgehammer include
relevance filters that heuristically select a subset of the available axioms;
filters such as MePo~\cite{meng-paulson-2009} and SInE
~\cite{hoder-voronkov-2011} are iterative and could be interleaved with
monomorphisation. Third, although one would expect native implementations of
polymorphism to outperform any preprocessor, this is not the case now, indicating that there is considerable room for improvement on the native front.

@@

   We thank Sascha Böhme for fruitful discussions and
   Jannis Limperg for suggesting textual improvements.

>>>>

Automatic theorem provers provide automation for proof assistant users. Many proof assistants, such as HOL4~\cite{slind-norrish-2008}, HOL Light~\cite{harrison-2009} and Isabelle\slash HOL \cite{nipkow-et-al-2002}, support rank-1 polymorphism, where type quantification is allowed at the top level of formulae. By contrast, many automatic provers operate only on monomorphic logics. One approach to closing this gap is to extend provers to natively support polymorphism, as has been done in Vampire~\cite{bhayat-reger-2020}. However, this entails a lot of work that must be redone for every prover.

@@

Any translation method based on instantiation will inevitably be incomplete.
By a type version of the compactness theorem, there exists for any first order polymorphic formula \(\varphi\) an equisatisfiable finite set of monomorphic instances of \(\varphi\). However, this set cannot be computed~\cite[Theorem~1]{expr-poly-types}.

B\"ohme's iterative approach is implemented as part of Isabelle/HOL's SMT (satisfiability modulo theories) integration~\cite[Chapter 2]{sb-phd}. This implementation is also used by Sledgehammer~\cite{judgement,paulson-blanchette-2010} to interface with superposition based automatic theorem provers. However, it is documented only as a single subsection in B\"ohme's PhD thesis~\cite[Section 2.2.1]{sb-phd}.
A similar algorithm appears to be implemented in the \texttt{MESON} tactic~\cite{harrison-1996} of HOL Light, but it is undocumented.

In this paper, we present an algorithm based on our understanding of Böhme's description and implementation (Section~\ref{sec:high level-algorithm}). We also provide a more detailed description to help future implementers. In addition, our description shows some of the ways in which an implementation can be made more efficient and avoid combinatorial explosions (Section~\ref{sec:low-level-algorithm}).

Briefly, the algorithm works as follows. We assume problems to be sets of formulae. All the problem's symbols are collected, and the polymorphic symbol instances are matched against the monomorphic ones. This yields new symbol instances, both polymorphic and monomorphic. The process is then iterated, making use of the newly generated instances.

Consider the unary type constructor \ty{list}. If a formula contains $\ty{list}(\alpha)$, where $\alpha$ is a type variable, the types $\ty{list}(\ty{int})$, $\ty{list}(\ty{list}(\ty{int}))$, etc., can be generated. However, because new types emerge through matching, $\ty{list}(\ty{list}(\ty{int}))$ can be obtained only once the $\ty{list}(\ty{int})$ instance has been generated.
This example also shows that an infinite set of instances can be generated.

To keep the number of generated formulae finite, we limit the number of iterations. After the iterations are completed, the new monomorphic symbol instances are used to instantiate the polymorphic symbols in the problem's formulae, generating new monomorphic formulae. Finally, because monomorphic provers support only nullary type constructors, types must be `mangled'; for example, the compound type $\ty{list}(\ty{int})$ might be mangled to $\ty{list\_int}$.

@@

The algorithm is sound because the newly generated formulae are instances of the initial problem's formulae. %, where type variables have been instantiated with monomorphic types.
However, it is not complete.

@@

This example illustrates how an infinite number of new formulae can be generated from a simple initial problem.
%Although the example may seem contrived, similar problems frequently arise in practice. For example, the \sym{concat} function of Isabelle/HOL, which is characterised in the base case by the equation \(\sym{concat}\langle\alpha\rangle\; \sym{Nil}\langle\ty{list}(\alpha)\rangle = \sym{Nil}\langle\alpha\rangle\), exhibits the same behavior.
/Any reasonable implementation must limit the number of new type arguments, substitutions and formulae.

@@

The algorithm presented above is too naïve for a practical implementation. In this section, we present a lower level algorithm with the following features.
First, numeric bounds are introduced to stop combinatorially explosive enumerations.
Second, type argument tuples are separated into an old set and a new set to avoid re-computing some of the same matchings in successive iterations.
Third, substitutions are directly applied to the type arguments instead of the formulae. This avoids having to re-extract the type arguments from the formulae at each iteration. New formulae are generated only once all iterations are completed, in a separate, final step.

The data structures used in the algorithm are based on the ones used in the high level description. Instead of a map \(M\) from symbols to monomorphic type argument tuples, we now have \(M_\text{old}\) and \(M_\text{new}\), which play the same role whilst also distinguishing between those type argument tuples that have already been matched against and those that have not. Similarly, \(N_\text{old}\) and \(N_\text{new}\) replace the map \(N\) from formulae to symbols to non-monomorphic type argument tuples. Finally, we maintain a map \(S\) from formulae to the substitutions generated by the matchings. It is used to generate new formulae in the final phase.

@@

The iterative monomorphisation algorithm is given in Figure~\ref{iter_mono}. It has three phases. The first phase applies a \emph{monomorphisation step} to each formula in \(\Phi\) until the user-set limit, \textcolor{ourblueviolet}{num\_loops}, is reached. This limit is the only bound necessary for the algorithm to terminate. We use the colour blue to identify bounds and code related to bounds. At the end of each of these iterations, the old and new type argument maps are updated with newly generated types. Once these iterations are completed, the first phase is complete and the substitutions used to create new type argument tuples are passed to $\mathit{generate\_mono\_formulae}$ for the second phase. The third phase mangles the composite types of the newly monomorphised formulae. This allows targeting a simply typed logic with no support for $n$-ary type constructors.
%~\cite[Sections 4 and 5]{mono-trans}.
% Paper name is gt

@@

Similar bounds are used for the non-monomorphic type argument tuples:
The limit on the number of new non-monomorphic type argument tuples is \(\min(\max(\PolyMult\cdot n, \PolyFloor),\allowbreak \PolyCap)\), where \(n\) is the number of non-monomorphic type argument tuples associated with the current formula. An important difference with the monomorphic case is that \(n\) depends on the current formula being processed whilst \(m\) does not.
Both in the monomorphic and in the non-monomorphic case, the maximum number of newly generated type argument tuples is fixed per formula and per iteration.

@@

                  \(\sigma\leftarrow\sigma_1\circ\dots\circ\sigma_n\)\strut

@@

The various bounds presented here overlap to some extent. For instance, having at most \textcolor{ourblueviolet}{substitution\_cap} substitutions generated by \emph{matches} may be sufficient to curb the number of new type argument tuples, making the \textcolor{ourblueviolet}{\MonoCap}, \textcolor{ourblueviolet}{\MonoMult}, \textcolor{ourblueviolet}{\MonoFloor} triplet superfluous. Nonetheless every bound has uses. For example, problems that lead to few successful matches but many type argument tuples may benefit from a limit on the number of new type argument tuples whilst problems for which substitution generation is more combinatorially explosive may benefit from a limit on the number of generated substitutions.

@@

The Zipperposition implementation of the \emph{mono\_substs} function uses a map from type variables to substitutions instead of a set to filter the relevant substitutions from \(S\) efficiently. The \textcolor{ourblueviolet}{max\_substs} bound exists for two main reasons:
\begin{enumerate}
   \item The iterative monomorphisation algorithm can generate up to \textcolor{ourblueviolet}{max\_substs} new monomorphic formulae per initial polymorphic formula. Generating an excessive number of new formulae can overwhelm the prover. The final number of output formulae is limited to at most \(|\Phi|\cdot \text{\textcolor{ourblueviolet}{max\_substs}}\).
   \item The \emph{mono\_substs} function is the algorithm's most combinatorially explosive part. For a formula \(\varphi\), if \(S(\varphi)\) contains \(n\) substitutions that each instantiate exactly one of \(v\) %different
type variables, up to \(n^v\) monomorphising substitutions may be generated. Recall that the total number of type argument tuples used to generate \(S(\varphi)\) can be doubly exponential in the number of loop iterations. The starting \(n\) may therefore already be very large.
\end{enumerate}
%[This function can be thought of as exploring the tree of all possible monomorphising substitutions, \(S_{\text{res}}\) would represent the leaves that have already been visited and \(\sigma\) would represent the current node of the tree.]

@@

Each bound of the monomorphisation process represents a tradeoff: a higher bound allows for a more exhaustive instantiation of type variables but takes more time. Since we cannot test all possible combinations of values for all bounds to find the best compromise between completeness and speed,
we group closely related bounds together and test combinations of values for the bounds in these groups. Once we find the best performing set of values for a group, we assign these values to the corresponding bounds as we begin the search for the next group. If several groups result in the same number of proved problems, we select the most constraining values. Winning entries are shown in bold in Tables \ref{mono_ty_args} to \ref{e_settings}.

@@

Table~\ref{mono_ty_args} groups bounds that control the maximum number of newly generated monomorphic type arguments per formula and per iteration. The limit on newly generated type arguments is determined by three components that form a natural group of bounds. The table shows that generating no new monomorphic type arguments seems to be the best approach. This result may seem counterintuitive, but it is possible to monomorphise formulae without generating monomorphic type arguments. This is because non-monomorphic type argument generation can generate substitutions that instantiate one or more type variables, and it is these substitutions that are used to monomorphise formulae.

@@

\looseness=-1
Table~\ref{nmon_ty_args} plays the same role as Table~\ref{mono_ty_args} except that it evaluates the bounds limiting the number of new \emph{non-}monomorphic type arguments. The bound values are lower because we found non-monomorphic type arguments to be combinatorially explosive in preliminary evaluations. The table confirms that non-monomorphic type argument generation drives the creation of useful non-monomorphic formulae. This is indicated by the very low number of problems solved when no new non-monomorphic type arguments are allowed. Performance of the monomorphisation algorithm seems to plateau for some ranges of values and drops off beyond.
%TODO check that 125 is what we get if we just run Zipperposition for 10 seconds (maybe with all additional problems that are easily monomorphised)

@@

For larger problems, the monomorphisation algorithm may time out despite the bounds. In these cases neither Zipperposition nor E will have had a chance to try to solve the problem. To avoid this, a timer can interrupt the monomorphisation algorithm, after which Zipperposition resumes normal operation. Table~\ref{mono_time} tests the amount of time that is allocated to monomorphisation against the number of iterations of the monomorphisation algorithm. Neither parameter seems to substantially affect the algorithm's performance.

@@

Table \ref{e_settings} shows the impact of the options with which we call the E prover. The call point corresponds to the fraction of the total timeout of Zipperposition's timeout after which the normal operation of Zipperposition is interrupted to begin the monomorphisation process. The E timeout limits how long E is run before being interrupted, at which point Zipperposition resumes normal operation. The longer Zipperposition runs before E is invoked, the more formulae are generated, and the more combinatorially explosive iterative monomorphisation is. This likely explains the poor performance for greater values of the E call point.

@@

To evaluate the usefulness of iterative monomorphisation as an alternative to native polymorphism, two competitive higher order polymorphic provers were run on the 1034 problem suite. We chose Leo-III and Zipperposition. Unfortunately we needed to exclude Vampire because of parsing issues.

@@

\looseness=-1
We see the following avenues for future work. First, iterative monomorphisation
blindly enumerates candidate instantiations, without exploiting any knowledge
about the logical structure of the formulae in which symbols occur. For
example, a lemma $\sym{p}\langle\alpha\rangle$ cannot be used to prove the
conjecture $\lnot \sym{p}\langle\ty{nat}\rangle$ because of the incompatible
polarities, but our algorithm instantiates $\alpha$ with $\ty{nat}$ regardless.
Second, some automatic provers as well as tools such as Sledgehammer include
relevance filters that heuristically select a subset of the available axioms;
filters such as MePo~\cite{meng-paulson-2009} and SInE~\cite{hoder-voronkov-2011} are iterative and could be interleaved with
monomorphisation. Third, although one would expect native implementations of
polymorphism to outperform any preprocessor, currently this is not the case, suggesting that there is considerable room for improvement on the native front.

@@

   We thank Sascha Böhme for fruitful discussions.
   We thank Jannis Limperg and Mark Summerfield for suggesting textual improvements.
